Uncertainty-aware graph-based multimodal remote sensing detection of out-of-distribution samples

### 2021-07-19 latex/CEUR/paper.tex
CDCEO submission sent.

% Used for KL-divergence
\usepackage{mathtools}
\DeclarePairedDelimiterX{\infdivx}[2]{(}{)}{%
  #1\;\delimsize\|\;#2%
}
\newcommand{\infdiv}{D\infdivx}
\DeclarePairedDelimiter{\norm}{\lVert}{\rVert}

@@

% The "title" command
\title{Uncertainty-aware graph-based multimodal remote sensing}

@@

\section{Uncertainty framework}
\label{sec::unc_framework}

@@

\subsection{Convergence assistance techniques}
In order to assist the convergence of subjective models, two additional assistance techniques have been used: teacher knowledge distillation and the use of a Dirichlet prior.
These have been shown to allow subjective models to provide better uncertainty estimates \cite{ZhaoXujiang2020}.
\subsubsection{Teacher knowledge distillation}
By training a non-subjective model in advance, its outputs, $\hat{p}_{ik}$, can be used in order to encourage the subjective model to converge to node Dirichlet distributions with $\mathbb{E}[p_{ik}]$ which are close to the teacher's deterministic estimates.
This is achieved using an additional term in the loss function,
\begin{equation}
    \mathcal{L}_{\text{T}}(\boldsymbol{\theta})=\sum_i\sum_k\left(\hat{p}_{ik}\log\frac{\hat{p}_{ik}}{\mathbb{E}[p_{ik}]}\right),
\end{equation}
which corresponds to the summation of KL-divergence terms between the teacher output probability and the expected value of the subjective model's Dirichlet distribution for each node, $\sum_i\infdiv{\hat{p}_{ik}}{\mathbb{E}[p_{ik}]}$.
Notice that this sum is computed over all nodes as opposed to just the nodes in $\mathbb{L}$.
Models trained using a teacher are denoted using the ``-T'' suffix e.g. a S-BGCN-T model would indicate that a pre-trained GCN was used as a teacher in order to assist the training convergence of a subjective graph convolutional model.
\subsubsection{Dirichlet prior}
A second convergence assistance technique which can be used involves the use of a Dirichlet prior, $\hat{\boldsymbol{\alpha}}$.
The exact method chosen to provide $\hat{\boldsymbol{\alpha}}$ will depend on the nature of the problem but we will assume nodes which are nearby in the graph are more likely to belong to the same output class than nodes which are far apart, a property known as homophily \cite{HuangQian2019}.
Using this assumption, we can use the computed distances on the graph to assign contributions of evidence from observed node labels to the other nodes in the graph using a function of our choosing.
If $d_{ij}$ denotes the shortest path distance between a given node, indexed by $i$ and an observed node, indexed by $j$, then the amount of evidence contributed to suggest that the $i^\text{th}$ node belongs to the $k^\text{th}$ class is given by
\begin{equation}
    h_{ik}(y_j,d_{ij})=
    \begin{cases}
      \frac{\exp{\left(\frac{-d_{ij}^2}{2\sigma^2}\right)}}{\left(2\pi\sigma^2\right)^{1/2}}, & \text{if } y_{jk}=1,\\
    0, & \text{otherwise},
    \end{cases}
\end{equation}
where $\sigma$ is a scale parameter which controls the order of distance magnitude over which evidence will propagate in the prior.
The total evidence to suggest the $i^\text{th}$ node belongs to the $k^\text{th}$ class, $e_{ik}$ can be found by summing these contributions over the nodes in the training set, such that the element in the prior is given by 
\begin{equation}
    \hat{\boldsymbol{\alpha}}_{ik}=1+e_{ik}=1+\sum_{j\in\mathbb{L}}h_{ik}(y_j, d_{ij}).
\end{equation}
The KL-divergence between the Dirichlet distribution of the prior and the model output is given by the term
\begin{equation}
    \mathcal{L}_{\text{K}}(\boldsymbol{\theta})=\sum_i\infdiv{\text{Dir}(\mathbf{p}_i\vert\boldsymbol{\alpha}_i)}{\text{Dir}(\hat{\mathbf{p}}_i\vert\hat{\boldsymbol{\alpha}}_i)},
\label{eq::teacher_loss}
\end{equation}
which can, in turn, be incorporated into the total loss function.
Models trained using a prior are denoted using the ``-K'' suffix.

@@

\section{Results and analysis}
\label{sec::results}
\subsection{Data}
A subsection of the 2018 IEEE GRSS Data Fusion Challenge dataset was selected for the purposes of validating the described methods.
The ground truth labels in this dataset describe 20 different urban land cover/land use classes (i.e. $K=20$) as well as an unlabelled state, described as Unclassified.
The modes of input data represent measurements from three sensor types: LiDAR, optical and hyperspectral (HS).
The LiDAR data was provided at $0.5$ m resolution, the same resolution as the ground truth labels (GT).
In order to simplify analysis, the optical data (which was provided at $0.05$ m resolution) and the HS data (which was provided at $1.0$ m resolution) were bilinearly resampled to obtain $0.5$ m resolution across inputs and outputs.

@@

In this paper we have adapted a novel classification method capable of providing uncertainty estimates to the task of multi-class classification of multimodal remote sensing data.
The adopted framework, based upon the theory of Subjective Logic, provides measures of vacuity and dissonance uncertainty.
Of the types of uncertainty assessed, the measure of vacuity was the best metric to perform identification of out of distribution samples.
Experimental results have shown the performance of the S-BGCN-T-K model for OOD detection to be improved against baseline methods.
This represents a promising avenue for uncertainty-aware learning in the task of multimodal remote sensing classification.

>>>>

% Used for KL-divergence
\usepackage{mathtools}
\DeclarePairedDelimiterX{\infdivx}[2]{(}{)}{%
  #1\;\delimsize\|\;#2%
}
\newcommand{\infdiv}{D_{\textrm{KL}}\infdivx}
\DeclarePairedDelimiter{\norm}{\lVert}{\rVert}

@@

% The "title" command
\title{Uncertainty-aware graph-based multimodal remote sensing detection of out-of-distribution samples}

@@

\section{Uncertainty framework}
\label{sec::unc_framework}
The proposed uncertainty-aware framework relies on the definition of uncertainty metrics, which in turn are based on subjective logic and a Dirichlet mapping \cite{ZhaoXujiang2020}.
These steps are detailed in this section, and have been properly adapted to the task of multimodal remote sensing classification.

@@

\subsection{Convergence assistance techniques}
In order to assist the convergence of subjective models, two additional assistance techniques have been used: teacher knowledge distillation and the use of a Dirichlet prior.
These have been shown to allow subjective models to provide better uncertainty estimates \cite{ZhaoXujiang2020}.
\subsubsection{Teacher knowledge distillation}
By training a non-subjective model in advance, its outputs, $\hat{p}_{ik}$, can be used in order to encourage the subjective model to converge to node Dirichlet distributions with $\mathbb{E}[p_{ik}]$ which are close to the teacher's deterministic estimates.
This is achieved using an additional term in the loss function,
\begin{equation}
    \mathcal{L}_{\text{T}}(\boldsymbol{\theta})=\sum_i\sum_k\left(\hat{p}_{ik}\log\frac{\hat{p}_{ik}}{\mathbb{E}[p_{ik}]}\right),
\end{equation}
which corresponds to the summation of Kullback-Leibler (KL) divergence terms between the teacher output probability and the expected value of the subjective model's Dirichlet distribution for each node.
Using $D_{\textrm{KL}}(\,\vert\vert\,)$ to compute the KL divergence, this is stated equivalently as $\sum_i\infdiv{\hat{p}_{ik}}{\mathbb{E}[p_{ik}]}$.
Notice that this sum is computed over all nodes as opposed to just the nodes in $\mathbb{L}$.
Models trained using a teacher are denoted using the ``-T'' suffix e.g. a S-BGCN-T model would indicate that a pre-trained GCN was used as a teacher in order to assist the training convergence of a subjective graph convolutional model.
\subsubsection{Dirichlet prior}
A second convergence assistance technique which can be used involves the use of a Dirichlet prior, $\hat{\boldsymbol{\alpha}}$.
The exact method chosen to provide $\hat{\boldsymbol{\alpha}}$ will depend on the nature of the problem but we will assume nodes which are nearby in the graph are more likely to belong to the same output class than nodes which are far apart, a property known as homophily \cite{HuangQian2019}.
Using this assumption, we can use the computed distances on the graph to assign contributions of evidence from observed node labels to the other nodes in the graph using a function of our choosing.
If $d_{ij}$ denotes the shortest path distance between a given node, indexed by $i$ and an observed node, indexed by $j$, then the amount of evidence contributed to suggest that the $i^\text{th}$ node belongs to the $k^\text{th}$ class is given by
\begin{equation}
    h_{ik}(y_j,d_{ij})=
    \begin{cases}
      \frac{\exp{\left(\frac{-d_{ij}^2}{2\sigma^2}\right)}}{\left(2\pi\sigma^2\right)^{1/2}}, & \text{if } y_{jk}=1,\\
    0, & \text{otherwise},
    \end{cases}
\end{equation}
where $\sigma$ is a scale parameter which controls the order of distance magnitude over which evidence will propagate in the prior.
The total evidence to suggest the $i^\text{th}$ node belongs to the $k^\text{th}$ class, $e_{ik}$ can be found by summing these contributions over the nodes in the training set, such that the element in the prior is given by 
\begin{equation}
    \hat{\boldsymbol{\alpha}}_{ik}=1+e_{ik}=1+\sum_{j\in\mathbb{L}}h_{ik}(y_j, d_{ij}).
\end{equation}
The KL divergence between the Dirichlet distribution of the prior and the model output is given by the term
\begin{equation}
    \mathcal{L}_{\text{K}}(\boldsymbol{\theta})=\sum_i\infdiv{\text{Dir}(\mathbf{p}_i\vert\boldsymbol{\alpha}_i)}{\text{Dir}(\hat{\mathbf{p}}_i\vert\hat{\boldsymbol{\alpha}}_i)},
\label{eq::teacher_loss}
\end{equation}
which can, in turn, be incorporated into the total loss function.
Models trained using a prior are denoted using the ``-K'' suffix.

@@

\section{Results and analysis}
\label{sec::results}
\subsection{Data}
A subsection of the 2018 IEEE GRSS Data Fusion Challenge dataset \cite{HoustonDataset} ws selected for the purposes of validating the described methods.
The ground truth labels in this dataset describe 20 different urban land cover/land use classes (i.e. $K=20$) as well as an unlabelled state, described as Unclassified.
The modes of input data represent measurements from three sensor types: LiDAR, optical and hyperspectral (HS).
The LiDAR data was provided at $0.5$ m resolution, the same resolution as the ground truth labels (GT).
In order to simplify analysis, the optical data (which was provided at $0.05$ m resolution) and the HS data (which was provided at $1.0$ m resolution) were bilinearly resampled to obtain $0.5$ m resolution across inputs and outputs.

@@

In this paper we have adapted a novel classification method capable of providing uncertainty estimates to the task of multi-class classification of multimodal remote sensing data.
The adopted framework, based upon the theory of Subjective Logic, provides measures of vacuity and dissonance uncertainty.
Of the types of uncertainty assessed, the measure of vacuity was the best metric to perform identification of OOD samples.
Experimental results have shown the performance of the S-BGCN-T-K model in the task of OOD detection to be improved against baseline methods.
This represents a promising avenue for uncertainty-aware learning in the task of multimodal remote sensing classification.
